Incorporating periodic variability in hidden Markov models for animal movement

### 2016-06-22 paper3.Rnw
working on paper

\documentclass{article}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Incorporating periodic variability in hidden Markov models for animal movement}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
probability that individuals occupy or move between groups can obscure
such signals. We use finite mixture and hidden Markov models, two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely explored model extension, resolves some shortcomings of current
modeling frameworks and clarifies the behavioural patterns of
panthers. More generally, incorporating previously neglected
structure in statistical models can allow researchers to more
accurately identify models of appropriate complexity.

@@

%% BMB: note that it is *much* easier to keep track of changes in Github,
%% which works on the line level, if you don't use line-wrap ...

Given a sequence of animal movements observed directly 
or via telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} %
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin1998quantitative}.

Ever-increasing capabilities of remote
sensors are making movement data available over an
ever-wider range of time scales, at both higher resolution (e.g. hourly data
from GPS collars vs. daily or weekly fixes for radio or VHF
collars) and longer extent (e.g. from a few days to significant
fractions of a year, or longer).  When analyzing such
remote-sensing data, ecologists will more
often have to account for temporal variability in movement
behaviour at diurnal and seasonal scales that were previously
not captured in the data.

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales \citep{gurarie2009novel}.

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which corresponds
to making precisely this assumption.

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).
(Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations while neglecting possible correlations
between them.)

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
\todo{it occurs to me that we haven't really included any
information about simulation testing in this version!
We should probably add a short section; if don't, then we
should delete following sentence}
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
There are several efficient algorithms for computing the marginal likelihood and
numerically estimating parameters \citep{zucchini_hidden_2009};
we use those implemented in the \texttt{depmixS4} package for R
\citep{visser2010depmixs4,R}.

@@

In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
As is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e. we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

>>>>

\documentclass{article}
\usepackage[sort]{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Incorporating periodic variability in hidden Markov models for animal movement}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

Clustering time-series data into discrete groups can improve
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers. More generally, we point out that model misspecification
(omitting important sources of variation) can lead to overfitting,
and as a corollary that incorporating previously neglected
structure in statistical models can allow more
accurate assessment of appropriate model complexity.

@@

\todo{ethics statement? numbers vs words?}
%% BMB: note that it is *much* easier to keep track of changes in Github,
%% which works on the line level, if you don't use line-wrap ...

Given a sequence of animal movements, movement models aim to find a parsimonious description
that can be used to understand past movements and predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} %
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour 
at particular times is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin1998quantitative}.

Ever-increasing capabilities of remote
sensors are making movement data available over an
ever-wider range of time scales, at both higher resolution (e.g. hourly data
from GPS collars vs. daily or weekly fixes for radio or VHF
collars) and longer extent (e.g. from a few days to significant
fractions of a year, or longer).  When analyzing such
long-term data, ecologists will more
often have to account for temporal variability in movement
behaviour at diurnal and seasonal scales that were previously
not captured in the data.

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
% BMB: pick one (perhaps include a parenthetical statement/definition
% saying that we treat these as a synonyms.  Just stringing them together
% with slashes is awkward)
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
direct comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales \citep{gurarie2009novel}.

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
As \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results on optimal
model complexity might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  For large data sets, 
information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not given), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which corresponds
to making precisely this assumption.

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).
(Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
\todo{it occurs to me that we haven't really included any
information about simulation testing in this version!
We should probably add a short section; if don't, then we
should delete following sentence}
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
There are several efficient algorithms for computing the marginal likelihood and
numerically estimating parameters \citep{zucchini_hidden_2009};
we used those implemented in the \texttt{depmixS4} package for R
\citep{visser2010depmixs4,R}.

@@

In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
As is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e. we have only $n \times (n-1)$ distinct parameters; we index $j$ from 2 to $n$ for notational clarity):
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}
